---
title: 'Lab 5: Anatomy of a Package and Package Documentation'
author: "Gregor Thomas"
date: "Monday, February 8, 2016"
output:
  ioslides_presentation:
    incremental: no
  slidy_presentation:
    fig_height: 3.5
    fig_width: 5
    highlights: pygments
    incremental: no
    theme: cosmo
---

## Anatomy of an R Package

R packages aren't all that special, just a collection of files. They include

- R files
- Documentation files
- DESCRIPTION and NAMESPACE files

in a common structure.

Nothing has to be compiled, the typical way of distributing package is just to compress the files together in a `.tar.gz` "tarball". This is called *bundling*.

Packages on CRAN are often compressed further into *binaries*, which are OS-specific, smaller to download, but not human-readable.

## Library vs Package

A *library* is a folder on your computer where you store packages. A *package* is a folder that has a specific structure so R knows how to make its functions and documentation available to you when you load it.

You can see libraries on your search path with `.libPaths()` (also used for adding a new library location).

```{r}
.libPaths()
```

## Inside the package folder

The package folder must have the same name as the package. Inside, four components are absolutely necessary:

- `DESCRIPTION` a text file (with no file extension) that has a very specific format for the package name, version, maintainer, author, and dependencies
- `R/` a folder containing `.R` files defining the functions in the package
- `man/` a folder containing `.Rd` documentation files
- `NAMESPACE` another text file with no file extension that has a list of the functions exported 

---

Other contents of a package are optional and depend on "extra" things your package might do. For your final projects, you'll also need:

- `data/` a folder with `.rdata` or `.rds` files for data sets or other non-function objects available in your pacakge.
- `inst/doc/` a folder for vignettes or other large, external documentation files

## Let's poke around

You can look at installed packages in your library. Some will be installed as binaries, which are compressed and not human-readable, but you can download the source on CRAN.

I like to view packages and code on the (unofficial) CRAN mirror on Github: <https://github.com/cran/>. It's nice because you can search for code inside a package.

---

Common things you'll see looking in packages:

- `src/` a folder for non-R code used by the package (often C or FORTRAN)
- `README` text file (no file extension) with a brief intro to the package
- `NEWS` describes changes between versions of the package (a well-maintained NEWS file is a hallmark of a really well-maintained package)

More details (and more options) available in [Hadley's online book on R Packages](http://r-pkgs.had.co.nz/).

## Working with packages {.smaller}

In this wonderful age, there are lots of tools that help with package building. `devtools` is the main workhorse, it creates the basic structure, builds the package (turn it into a tarball), runs tests on it, writes your NAMESPACE file, etc.

`roxygen2` is even more useful. If you comment your functions in the roxygen way, the comments will be converted into R documentation. R documentation is written in a sort of lite version of LaTeX, and it's *very* nice to not do edit it manually.

Roxygen also keeps your NAMESPACE file nice and clean.

RStudion can create projects for a specific package you're working on that has some nice point-and-cick tools to use `devtools` functions.

## Documentation

```{r, echo = FALSE, include=FALSE}
packages = c("ggplot2", "roxygen2", "devtools")
lapply(packages, library, character.only = T)
theme_set(theme_bw(10))
```

Documentation is extremely important, both for recording your thoughts/decisions/conclusions as you conduct an analysis, and in your code so you can make sense of what you've written weeks, months, and years later.

> **Your most important research partners are the ones most difficult to communicate with:** your past self and your future self.

- The best you can do is leave yourself good notes.

## Function Documentation

Your homework, and especially your final project, are good practice for a narrative documenting an analysis. When you build a package, the main point is to bundle together functions that will be useful again for you or another user. Functions are useless to someone who doesn't understand what they do.

---

You've read countless R help files by this point, how much can you generalize about the structure of an R help file?

- title
- description
- usage
- arguments
- details
- value
- examples
- others references, citations, notes, warnings, see also, aesthetics (for ggplot2 functions)...       

How do you think they stack up in terms of importance?

## Rd files

R documentation has it's own file type, `.Rd`. They're written in a sort of LaTeX-lite markup language that is parsed to HTML when you view the documentation.

In the olden days (pre-2008), the .Rd files were written by hand, which is a real pain in terms of organization. Roxygen2 let's you put comments in your code, and it converts the comments to `.Rd` files for you. However, it still uses a lot of .Rd syntax, which means there's a lot of curly braces and backslashes. For example to emphasize text, you write `\emph{my text}`, and to do a monospace font (for code), you write `\code{my_code}`.

Roxygen doesn't read all your comments, it only reads comments that are marked with a single hash followed by a single quote, `#'`. It will associate the comment with R code that follows immediately.

## A quick example

A utility function I like is `samp`, it gives a quick random sample of rows from a data frame. I use it to check data as I clean it.

```{r}
samp <- function(x, n = 10) {
    if (n >= nrow(x)) return(x)
    return(x[sample(1:nrow(x), size = n), ])
}
```

--- 

Roxygen makes some assumptions (unless you tell it otherwise):
the first paragraph (or *block*) of comments is the title, the next
block (separated by a blank `#'` roxygen comment line) is the description

```{r}
#' Quick random sample of a data frame
#'
#' \code{samp} can be used to quickly check on a data frame.
samp <- function(x, n = 10) {
    if (n >= nrow(x)) return(x)
    return(x[sample(1:nrow(x), size = n), ])
}
```

---

After that, we use roxygen tags which put things in sections, notated `@section`. Function arguments are a little special, each should have its own line, they are tagged as `@param`s:

```{r}
#' Quick random sample of a data frame
#'
#' \code{samp} can be used to quickly check on a data frame.
#' @param x data frame to sample from
#' @param n positive integer number of rows to return in the sample
samp <- function(x, n = 10) {
    if (n >= nrow(x)) return(x)
    return(x[sample(1:nrow(x), size = n), ])
}
```

---

It's also good practice to describe what type of object is returned (if any), and to show an example (though this function is so simple it's getting over-documented). 

**N.B.** the plural `@examples` tag is used for the section. If singular, `@example`, roxygen will look for example code in a different file (sometimes used for lengthier examples).

```{r}
#' Quick random sample of a data frame
#' 
#' \code{samp} can be used to quickly check on a data frame.
#' @param x data frame to sample from
#' @param n positive integer number of rows to return in the sample
#' @return Returns a data frame
#' @examples
#' samp(mtcars)
samp <- function(x, n = 10) {
    if (n >= nrow(x)) return(x)
    return(x[sample(1:nrow(x), size = n), ])
}
```

---

If the function is to be **visible** to users (i.e., not an internal-only function), you add an `@export` tag. If the function depends on any non-base functions, you should also mark the dependencies with an `@import` tag (for a full package) or `@importFrom` (for a single imported function).

```{r}
#' Quick random sample of a data frame
#'
#' \code{samp} can be used to quickly check on a data frame.
#' @param x data frame to sample from
#' @param n positive integer number of rows to return in the sample
#' @return Returns a data frame
#' @examples
#' samp(mtcars)
#' @export
samp <- function(x, n = 10) {
    if (n >= nrow(x)) return(x)
    return(x[sample(1:nrow(x), size = n), ])
}
```

---

If the function depends on any non-base functions, you should also mark the dependencies with an `@import` tag (for a full package) or `@importFrom` (for a single imported function)... doesn't apply here.

```{r}
#' Quick random sample of a data frame
#'
#' \code{samp} can be used to quickly check on a data frame.
#' @param x data frame to sample from
#' @param n positive integer number of rows to return in the sample
#' @return Returns a data frame
#' @examples
#' samp(mtcars)
#' @export
samp <- function(x, n = 10) {
    if (n >= nrow(x)) return(x)
    return(x[sample(1:nrow(x), size = n), ])
}
```

## In action:

Let's put this in a package!

```{r, eval = FALSE}
library("devtools")
create("demo1")
# save the above code in demo1/R/samp.R

# then create the documentation
devtools::document("demo1")
```

## In action:

We can look at the .Rd file, and (in RStudio) preview the HTML help.
The header and the Usage section are generated automatically. Or you can run `devtools::load_all("demo1")` to simulate loading the package without really installing it.

## Your turn

Write roxygen comments for this simple function:

```{r}
abs_diff <- function(a, b) {
    abs(b - a)
}
```

and, if you've been following along, save your comments and the function as `demo1/R/abs_diff.R`, re-roxygenize, and view the results!

---

Quick clean-up, deleting the package we just created.

```{r, eval = FALSE}
unlink("demo1", recursive = T)
```

 Note that if you do this, you'll have to either unload the package or restart R before using help again, because it knows to try to look for help in this package's directory and it won't like that the directory is gone.

Unloading a package can be done like this:

```{r, eval = FALSE}
detach("package:demo1", unload=TRUE) # base R way
devtools::unload(pkg = "demo1")      # devtools way
```

## Roxygen Resources

Ironically, roxygen2 used to be very poorly documented. The 4.0 release changed that, and it now has some pretty good vignettes, viewable [on CRAN](http://cran.r-project.org/web/packages/roxygen2/index.html) or with `vignette(package = "roxygen2")`. 

If you start with `vignette(topic = "roxygen2")`, it links to the others. The "Generating Rd Files" is an expanded version of what we've covered here. In the others you can find details about all the section tags, the options for formatting text, a quick introduction to the NAMESPACE file.
