---
title: "Lab 3 Live"
author: "Gregor Thomas"
date: "January 25, 2016"
output: html_document
---

Load some packages

```{r, message = FALSE, results = "hide"}
packages = c("ggplot2", "arm", "broom", "reshape2", "tidyr", "dplyr")
lapply(packages, library, character.only = T)
```

Read in data and fit a model:

```{r}
wells = read.table("http://www.stat.columbia.edu/~gelman/arm/examples/arsenic/wells.dat")

wells$dist.log = log(wells$dist)

m_logd = glm(switch ~ dist.log, family = binomial(link = "logit"), data = wells)

display(m_logd)
```

Plot similar to Fig. 5.9

```{r}
p = ggplot(wells, aes(x = dist.log, y = switch)) +
    geom_point(position = position_jitter(height = 0.05),
               size = 0.8, alpha = 0.4) +
    stat_function(fun = function(x) {
        invlogit(coef(m_logd)[1] + coef(m_logd)[2] * x)
        })
```

```{r}
well_a = augment(m_logd)
ggplot(well_a, aes(x = dist.log, y = .resid)) +
    geom_point(position = position_jitter(height = 0.1), size = 0.8, alpha = 0.4)
```

Binned residual plot

```{r}
well_bin = well_a %>%
    mutate(
        q_group = cut(dist.log,
                      breaks = quantile(dist.log, seq(0, 1, length.out = 30), include.lowest = TRUE))
    ) %>%
    group_by(q_group) %>%
    summarize(.resid = mean(.resid),
              dist.log = mean(dist.log))

ggplot(well_bin, aes(x = dist.log, y = .resid)) +
    geom_point() +
    geom_smooth()

```


```{r}
wells$dist100 = wells$dist / 100
mod = glm(switch ~ dist100 + log(arsenic), data = wells,
          family = binomial)
display(mod)
```


```{r}
fill_in = function(x, n = 50) {
    seq(from = min(x), to = max(x), length.out = n)
}

new_well = expand.grid(
    dist100 = fill_in(wells$dist100),
    arsenic = exp(fill_in(log(wells$arsenic)))
)

nw_plot = ggplot(new_well, aes(x = dist100, y = arsenic)) +
    geom_point(size = 0.3)

nw_plot
nw_plot + scale_y_continuous(trans = "log")
```

```{r}
pred = augment(mod, newdata = new_well, type.predict = "response")

ggplot(pred, aes(x = dist100, y = arsenic)) +
    geom_tile(aes(fill = .fitted)) +
    stat_contour(aes(z = .fitted), colour = "black") +
    scale_y_continuous(trans = "log") +
    geom_point(data = wells, aes(colour = factor(switch)), alpha =  0.5) +
    scale_color_manual(values = c("white", "green"))
```

To make the "confusion matrix" faceted plot we need to turn the model's probability-scale predictions into binary 1-0 prdictions. Because we're not doing any fancy adjustments, we'll just round the fitted values to the nearest 1 or 0.

```{r}
well_a2 = augment(mod, type.predict = "response")
well_a2$predicted = round(well_a2$.fitted)

# this is the 'confusion matrix'
# 0,0 and 1,1 are the correct predictions
# the 0,1 and 1,0 are the incorrect predictions
table(well_a2$switch, well_a2$predicted)
```

The "true" values are in the `switch` column, and the predicted values are in the cleverly named `predicted` column. For our confusion matrix plot we need to facet by these two variables:


```{r}
ggplot(pred, aes(x = dist100, y = arsenic)) +
   scale_y_continuous(trans = "log") +
    geom_tile(aes(fill = .fitted)) +
        stat_contour(aes(z = .fitted), colour = "black") +
    geom_point(data = well_a2,
               aes(colour = factor(switch == predicted,
                                   labels = c("Incorrect", "Correct")),
                   y = exp(log.arsenic.)),
               alpha =  0.5) +
    scale_color_manual(values = c("firebrick3", "green"),
                       name = "Model Performance") +
    facet_grid(switch ~ predicted)
```


## Regular Expressions

Some sample strings:

```{r}
library(stringr)
counties = c("Adams", "Asotin", "Benton", "Chelan", "Clallam", "Clark",
"Columbia", "Cowlitz", "Douglas", "Ferry", "Franklin", "Garfield",
"Grant", "Grays Harbor", "Island", "Jefferson", "King", "Kitsap",
"Kittitas", "Klickitat", "Lewis", "Lincoln", "Mason", "Okanogan",
"Pacific", "Pend Oreille", "Pierce", "San Juan", "Skagit", "Skamania",
"Snohomish", "Spokane", "Stevens", "Thurston", "Wahkiakum", "Walla Walla",
"Whatcom", "Whitman", "Yakima")
```

`str_detect` returns true or false based on whether it finds the pattern in each string.

```{r}
# which counties have a lowercase 'a'?
str_detect(counties, pattern = "a")

# the logical vector isn't super useful, instead let's 
# subset the vector by the matches
counties[str_detect(counties, pattern = "a")]

# counties that have an 'a' followed by an 's' 
counties[str_detect(counties, pattern = "as")]

# a really long pattern (case matters!):
counties[str_detect(counties, pattern = "Skagi")]
```

`str_replace` is the *find/replace* command. Note that I am **not** assigning the results so as to leave the original vector intact. 

```{r}
# replace 'a' with space ' '
str_replace(counties, pattern = "a", replacement = " ")

# str_replace only replaces the *first* instance per string
# to get them all, use str_replace_all
str_replace_all(counties, pattern = "a", replacement = " ")

str_replace(counties, pattern = "fer", replacement = " FER was here ")

# replace all spaces with nothing (delete spaces)
str_replace_all(counties, pattern = " ", replacement = "")
```

Special characters! There are lots of special characters, [this cheatsheet](http://www.cheatography.com/davechild/cheat-sheets/regular-expressions/) is helpful.

```{r}
# '^' is the start of the string
# '^A' will match strings that start with 'A'
counties[str_detect(counties, pattern = "^A")]

# '$' is the end of the string
# 's$' matches strings that end with s
counties[str_detect(counties, pattern = "s$")]

# '.' matches any single character
counties[str_detect(counties, pattern = "o.o")]

# First four characters:
str_extract(counties, pattern = "^....")

# You can use quantifier in curly braces
# (this is the same as the line above)
str_extract(counties, pattern = "^.{4}")

# and of course we can mix and match,
# any 4 characters followed by an n in the 5th position
str_extract(counties, pattern = "^.{4}n")
```

Escaping special characters: what if you actually want to match a period?
In regex, all of the special characters have special meaning. If you want to match them, you need to *escape* them with **two** backslashes:

```{r}
# replaces all characters with nothing (probably not what you want)
str_replace_all(names(well_a), pattern = ".", "")

# replaces periods with nothing (works because we escaped the period)
str_replace_all(names(well_a), pattern = "\\.", "")

# Quotes can also be escaped
x = "Escaping \"quotes\" inside a \'string\' works with a single backslash."
y = "backslashes \\ must also be escaped"
z = "Use the cat command to see what's \"really there\""

print(x)
print(y)
print(z)
cat(x)
cat(y)
cat(z)

# you can also alternate single and double quotes
# but this can only go one level deep:
a = "This string has 'quotes' in it."
b = 'This string has "double quotes" in it.'
```

"Just keep adding adding backslashes until it works"


## For loops


See the new For Loop Discussion Forum!
